% Required packages: amsmath, amssymb

\section{Grid-Free Geometry-Interior MAF}
\label{sec:geometry_calibrated_maf}

MAF combines the confidence score $s_{\mathrm{conf}}(x)$ and the consistency
score $s_{\mathrm{cons}}(x)$ by
\begin{equation}
    s_{\alpha}(x)
    =
    s_{\mathrm{conf}}(x)^{\alpha}
    s_{\mathrm{cons}}(x)^{1-\alpha},
    \qquad
    \alpha \in [0,1].
\end{equation}
Previous experiments selected $\alpha$ by sweeping a finite grid such as
$\{0.0,0.1,\ldots,1.0\}$.  This is undesirable because the selected value is
not a property of the ID model alone; it depends on the OOD evaluation set used
for model selection.  We therefore replace grid search by an ID-only geometric
rule that returns a single deterministic value of $\alpha$.

\subsection{ID-only Leave-one-class-out Geometry}

Let $h(x)\in\mathbb{R}^{d}$ be the frozen backbone feature.  For each ID class
$c$, we construct a leave-one-class-out fold.  The classes
$\mathcal{C}\setminus\{c\}$ are treated as ID, while class $c$ is treated as a
pseudo-OOD class.  On this fold, class means and a tied covariance are estimated
only from $\mathcal{C}\setminus\{c\}$, and Mahalanobis distances produce
fold-specific scores
$s_{\mathrm{conf}}^{(-c)}(x)$ and $s_{\mathrm{cons}}^{(-c)}(x)$.
Each sample is then mapped to the two-dimensional log-score geometry
\begin{equation}
    r^{(-c)}(x)
    =
    \begin{pmatrix}
        \log(s_{\mathrm{conf}}^{(-c)}(x)+\epsilon) \\
        \log(s_{\mathrm{cons}}^{(-c)}(x)+\epsilon)
    \end{pmatrix}
    \in \mathbb{R}^{2}.
\end{equation}
Collecting all folds gives an ID set $R_{+}$ and a pseudo-OOD set $R_{-}$:
\begin{align}
    R_{+}
    &=
    \bigcup_{c\in\mathcal{C}}
    \{r^{(-c)}(x_i): y_i\neq c\},\\
    R_{-}
    &=
    \bigcup_{c\in\mathcal{C}}
    \{r^{(-c)}(x_i): y_i=c\}.
\end{align}
Let
\begin{equation}
    \delta = m_{+}-m_{-},
    \qquad
    S = \operatorname{Cov}(R_{+})+\operatorname{Cov}(R_{-})+\lambda I_{2},
\end{equation}
where $m_{+}$ and $m_{-}$ are the two empirical means and $\lambda>0$ is a small
numerical ridge.

\subsection{Geometry-Interior Shrunk Fisher Alpha}

In log space,
\begin{equation}
    \log s_{\alpha}(x)
    \simeq
    w_{\alpha}^{\top} r(x),
    \qquad
    w_{\alpha}
    =
    \begin{pmatrix}
        \alpha\\
        1-\alpha
    \end{pmatrix}.
\end{equation}
The unconstrained Fisher direction for separating $R_{+}$ from $R_{-}$ is
\begin{equation}
    a = S^{-1}\delta
    =
    \begin{pmatrix}
        a_{\mathrm{conf}}\\
        a_{\mathrm{cons}}
    \end{pmatrix}.
\end{equation}
If both components of $a$ are positive, the Fisher direction lies inside the
nonnegative MAF cone and can be normalized to the MAF simplex:
\begin{equation}
    \hat{\alpha}_{F}
    =
    \frac{a_{\mathrm{conf}}}
         {a_{\mathrm{conf}}+a_{\mathrm{cons}}}.
\end{equation}
However, a valid cone direction is not always a reliable final mixture.  The
DINOv2 case shows a different failure mode: the Fisher direction remains in the
positive cone, but its normalized value saturates near the simplex boundary
($\hat{\alpha}_{F}\simeq 1$), whereas the OOD-tuned oracle is near
$\alpha=0.5$.  Thus the method should penalize both out-of-cone directions and
near-boundary directions without introducing a new OOD-tuned threshold.

We first define the conservative cone-corrected Fisher value
\begin{equation}
    \hat{\alpha}_{\mathrm{CIF}}
    =
    \begin{cases}
    \displaystyle
    \frac{a_{\mathrm{conf}}}
         {a_{\mathrm{conf}}+a_{\mathrm{cons}}},
    & \text{if } a_{\mathrm{conf}}>0
      \text{ and } a_{\mathrm{cons}}>0,\\[2.2ex]
    \displaystyle
    \frac{1}{2},
    & \text{otherwise.}
    \end{cases}
    \label{eq:cif_alpha}
\end{equation}
The fallback $\alpha=1/2$ is not tuned on OOD data.  It is the maximum-entropy
point of the two-component MAF simplex and corresponds to the neutral product
score
\begin{equation}
    s_{1/2}(x)
    =
    \sqrt{s_{\mathrm{conf}}(x)s_{\mathrm{cons}}(x)}.
\end{equation}

To handle cone-valid but boundary-saturated directions, we attach an
interiorness weight to the Fisher displacement from the neutral point:
\begin{equation}
    \gamma_F
    =
    \begin{cases}
    4\hat{\alpha}_{F}(1-\hat{\alpha}_{F}),
    & \text{if } a_{\mathrm{conf}}>0
      \text{ and } a_{\mathrm{cons}}>0,\\
    0,
    & \text{otherwise.}
    \end{cases}
\end{equation}
This weight is one at the simplex center and approaches zero at either
endpoint.  The proposed geometry-interior shrunk value is
\begin{equation}
    \hat{\alpha}_{\mathrm{GIS}}
    =
    \frac{1}{2}
    +
    \gamma_F
    \left(\hat{\alpha}_{\mathrm{CIF}}-\frac{1}{2}\right).
    \label{eq:gis_alpha}
\end{equation}
Equivalently, for a valid Fisher direction,
$\gamma_F=1-|2\hat{\alpha}_{F}-1|^2$.  Hence no threshold is introduced:
boundary saturation continuously reduces the amount of Fisher displacement.
For example, $\hat{\alpha}_{F}=0.95$ gives
$\gamma_F=4(0.95)(0.05)=0.19$ and
$\hat{\alpha}_{\mathrm{GIS}}\simeq 0.586$.

\paragraph{Final score.}
Given $\hat{\alpha}_{\mathrm{GIS}}$, the proposed score is
\begin{equation}
    s_{\mathrm{GIS\text{-}MAF}}(x)
    =
    s_{\mathrm{conf}}(x)^{\hat{\alpha}_{\mathrm{GIS}}}
    s_{\mathrm{cons}}(x)^{1-\hat{\alpha}_{\mathrm{GIS}}}.
\end{equation}
No OOD labels, OOD validation set, or grid search over $\alpha$ is used.

\subsection{Uniqueness and Diagnostic Quantities}

For fixed ID validation data and $\lambda>0$, $S$ is positive definite and
$a=S^{-1}\delta$ is unique.  Equations~\eqref{eq:cif_alpha}
and~\eqref{eq:gis_alpha} therefore return a single value of $\alpha$ for a
given backbone and ID validation split.  This
turns $\alpha$ from a swept hyperparameter into a deterministic geometric
quantity.

We also report three diagnostics:
\begin{align}
    \eta_{\mathrm{cone}}
    &=
    \mathbf{1}\{a_{\mathrm{conf}}>0,\;a_{\mathrm{cons}}>0\},\\
    \eta_{\mathrm{sat}}
    &=
    |2\hat{\alpha}_{F}-1|,\\
    \gamma_F
    &=
    4\hat{\alpha}_{F}(1-\hat{\alpha}_{F})
    \quad \text{when } \eta_{\mathrm{cone}}=1.
\end{align}
$\eta_{\mathrm{cone}}=0$ indicates that the Fisher direction is outside the MAF
simplex cone.  A large $\eta_{\mathrm{sat}}$ indicates that the ID Fisher rule
nearly collapses to a single component.  A small $\gamma_F$ means that the
final score will move only weakly away from the neutral product mixture.  These
diagnostics are useful for identifying backbones such as DINOv2, where ID
pseudo-OOD geometry and real OOD geometry may not be aligned.

\subsection{Difference from Oracle Alpha}

The purpose of $\hat{\alpha}_{\mathrm{GIS}}$ is not to reproduce the
OOD-tuned oracle value
\begin{equation}
    \alpha_{\mathrm{oracle}}
    =
    \arg\max_{\alpha\in[0,1]}
    \operatorname{AUROC}_{\mathrm{test\ OOD}}(\alpha).
\end{equation}
Instead, the purpose is to remove OOD-label tuning.  The oracle value is used
only for analysis.  The proposed method answers a different question:
given only ID validation geometry, which confidence--consistency mixture is
geometrically justified?

\subsection{Algorithm}

\begin{enumerate}
    \item Extract ID validation features $h(x_i)$ and labels $y_i$.
    \item For each class $c$, build a leave-one-class-out Mahalanobis model on
          $\mathcal{C}\setminus\{c\}$.
    \item Map validation samples to
          $r^{(-c)}(x)=(\log s_{\mathrm{conf}}^{(-c)}(x),
          \log s_{\mathrm{cons}}^{(-c)}(x))^{\top}$.
    \item Form $R_{+}$ from non-held-out samples and $R_{-}$ from held-out
          samples.
    \item Compute $\delta=m_{+}-m_{-}$ and
          $S=\operatorname{Cov}(R_{+})+\operatorname{Cov}(R_{-})+\lambda I_2$.
    \item Compute $a=S^{-1}\delta$.
    \item If $a_{\mathrm{conf}}>0$ and $a_{\mathrm{cons}}>0$, set
          $\hat{\alpha}_{\mathrm{CIF}}=
          a_{\mathrm{conf}}/(a_{\mathrm{conf}}+a_{\mathrm{cons}})$.
          Otherwise set $\hat{\alpha}_{\mathrm{CIF}}=1/2$.
    \item Set
          $\gamma_F=4\hat{\alpha}_{\mathrm{CIF}}(1-\hat{\alpha}_{\mathrm{CIF}})$
          if the Fisher direction is cone-valid, and $\gamma_F=0$ otherwise.
          Then compute
          $\hat{\alpha}_{\mathrm{GIS}}
          =1/2+\gamma_F(\hat{\alpha}_{\mathrm{CIF}}-1/2)$.
    \item Score test samples by
          $s_{\mathrm{GIS\text{-}MAF}}(x)$.
\end{enumerate}
